A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-16 docs/methods/erosion_methods.tex
NM
addressed Nick's todo comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform the above simulation, two fast and accurate methods are employed---one to solve the fluid equations and the other to evolve the eroding grains.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we employ a double-layer potential so that the resulting BIE is second-kind.  This second-kind BIE is discretized with a spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function. The linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is determined, we compute the shear stress with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We consider flows that are confined by a solid wall $\Gamma$ that encloses $M$ eroding bodies as diagrammed in Fig.~\ref{fig:schematic}.  The bodies are denoted as $\gamma_\ell$, $\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup \gamma_M$.  On each body, $\nn$ and $\ss$ denote the unit normal and unit tangent vectors respectively.  Throughout the paper, we adopt the convention that $\ss$ points in the counterclockwise direction and $\nn$ points out of the fluid domain (i.e.~into the bodies). Neglecting inertial forces, the dynamics of the fluid is fully determined by the position of the bodies $\xx_\ell(s,t) \in \gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given $\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the bounding wall, we prescribe a velocity $\UU(\xx,t)$.  Then, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{\em conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{\em no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{\em outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{\em erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the fluid viscosity.  The constant of proportionality of the erosion law only sets the time scale, so we have set $\Vn = |\tau|$ as described in equation~\eqref{eqn:erosionModel}.  
\todo[inline]{The above sentence is awkard. Could it be rephrased? First, the reader probably does not know what constant of proportionality you are referring to because we do not have one in the equation. Second, there are too many 'ofs' in a row.}
For the sake of numerical stability, we will modify this law in Section~\ref{sec:thetaL}.

@@

\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing equations.  A no-slip boundary condition is imposed on each body $\gamma_\ell$ whose unit normal points outward relative to the geometry.  On the outer geometry $\Gamma$, a Hagen-Poiseuille flow is imposed.  The bodies are eroded at a rate that is proportional to the magnitude of the shear stress applied by the Stokesian fluid.  The bodies are constrained to the middle third of the channel that is located between the dashed lines.}
\end{figure}
\todo[inline]{Here is an example where you can avoid the passive voice because it is totally unnecessary. Instead of saying 'the bodies are eroded', simply say 'the bodies erode'. Also, a fluid cannot be Stokesian, since the Reynolds number depends on the length scale and velocity scale. It is the flow that is Stokes.}

@@

In this section, we describe an equivalent BIE formulation of the incompressible Stokes equations with Dirichlet boundary conditions
\begin{equation}
  \begin{split}
  \mu\Delta \uu &= \grad p\, , \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
  \end{split}
  \label{eqn:stokes}
\end{equation}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on $\Gamma$, and it is zero on the eroding grains $\gamma$.  In two dimensions, a BIE for a stream function formulation is possible~\cite{gre-kro-may1996}, but this does not extend to three dimensions.  Therefore, we adopt a primitive variable formulation where the velocity is written in terms of the standard double-layer potential~\cite{lad1963,poz1992}
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy\, , \quad \xx \in \Omega \, ,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown density function.  As the viscosity $\mu$ also sets the time scale, we take $\mu=1$.

@@

This BIE formulation requires that $\bd\Omega$ is smooth.  Unfortunately, smoothness is not guaranteed by erosion since corners develop at locations where the shear stress vanishes.  One option is to modify our BIE to handle corners~\cite{rac-ser2017, ser-rok2016, hel2011, gil-hao-mar2014, bre2012}, but these methods require the corner's location, a non-uniform spaced mesh, or are ill-conditioned.  Instead, make a slight modification to the erosion model in Section~\ref{sec:thetaL} which allows us to control the smoothness of the grains.  
\todo[inline]{There is a typo in the above sentence (missing 'we'?). Also, technically there is  a dangling participle; i.e.~is it the erosion model which is in Section 2.3 or the modification? A slight rephrasing might clarify.}
In this manner, the BIE as described above can be applied.

@@

To numerically compute the completed double-layer potential~\eqref{eqn:completed_BIE}, we discretize each of the $M$ bodies with  $N_\iin$ points, and the outer boundary with $N_\out$ points.  Then, $N = MN_\iin + N_\out$ is the total number of discretization points of $\bd\Omega$.  The double-layer potential in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}\, , \quad i=1,\ldots,N \, ,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} \, ,
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is replaced with the limiting value of the double-layer potential
\begin{align}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)) \, ,
\end{align}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent vector at $\xx \in \bd\Omega$.  The equations for the Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are also discretized with the trapezoid rule.  Applying this discretization strategy to~\eqref{eqn:completed_BIE}, the dense linear system that solved is
\todo[inline]{Typo above}
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j) \, , \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j \, , \\ 
    \xi_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j \, ,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  We have applied the trapezoid rule since all the integrands are smooth and periodic, and this guarantees that the numerical solutions converges to the true solution with spectral accuracy~\cite{tre-wei2014}.
\todo[inline]{The above sentence strikes me as awkward and verbose. Rephrase? \\
How about this: Since all integrands are smooth and periodic, the trapezoid rule employed above achieves spectral accuracy.}

@@

We do not employ an FMM to evaluate the shear stress at $\xx_i$, $i=1,\ldots,N$, so the cost of evaluating the shear stress is $\bigO(N^2)$.  However, this calculation is only performed once per time step as opposed to the velocity double-layer potential, where we do employ the FMM, that must be evaluated at each GMRES iteration. 
\todo[inline]{The above sentence is confusing. Try to rephrase and shorten it.}
In future work, we plan to use a kernel-independent FMM~\cite{yin-bir-zor2004} to evaluate the shear stress with $\bigO(N)$ operations.

@@

For the Stokes equations, on boundaries that have a no-slip boundary condition, the shear stress is equivalent to the vorticity.  Therefore, the vorticity is computed on the grains $\gamma_\ell$ by numerically compute the shear stress as described in Section~\ref{sec:shearStress}.

@@

Naturally, the flow transmitted by the medium is tied to the amount of drag the bodies exert. We therefore show in Fig.~\ref{fig:mbodyplots}(c) the total drag exerted by all bodies as it varies in time. In both cases, the drag remains nearly constant throughout much of the simulation and then drops to zero in the final moments. Although all bodies are diminishing in size, the associated increase in $\umax$ leads to the nearly constant drag that is observed here.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform the above simulation, two fast and accurate methods are
employed---one to solve the fluid equations and the other to evolve the
eroding bodies.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we employ a double-layer potential so that the resulting BIE is second-kind.  This second-kind BIE is discretized with a spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function. The linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is determined, we compute the shear stress with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We consider flows that are confined by a solid wall $\Gamma$ that encloses $M$ eroding bodies as diagrammed in Fig.~\ref{fig:schematic}.  The bodies are denoted as $\gamma_\ell$, $\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup \gamma_M$.  On each body, $\nn$ and $\ss$ denote the unit normal and unit tangent vectors respectively.  Throughout the paper, we adopt the convention that $\ss$ points in the counterclockwise direction and $\nn$ points out of the fluid domain (i.e.~into the bodies). Neglecting inertial forces, the dynamics of the fluid is fully determined by the position of the bodies $\xx_\ell(s,t) \in \gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given $\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the bounding wall, we prescribe a velocity $\UU(\xx,t)$.  Then, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{\em conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{\em no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{\em outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{\em erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the fluid viscosity.  A scaling constant often appears in the erosion model in~\eqref{eqn:erosionModel}, but in the Stokes limit, the constant only sets the time scale, so we have taken it to be one.  For the sake of numerical stability, we will modify the erosion model in Section~\ref{sec:thetaL}.

%The constant of proportionality of the erosion law only sets the time scale, so we have set $\Vn = |\tau|$ as described in equation~\eqref{eqn:erosionModel}.  
%\todo[inline]{The above sentence is awkard. Could it be rephrased? First, the reader probably does not know what constant of proportionality you are referring to because we do not have one in the equation. Second, there are too many 'ofs' in a row.}

@@

\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing equations.  A no-slip boundary condition is imposed on each body $\gamma_\ell$ whose unit normal points outward relative to the geometry.  On the outer geometry $\Gamma$, a Hagen-Poiseuille flow is imposed.  The bodies erode at a rate that is proportional to the magnitude of the shear stress applied by the fluid.  The bodies are constrained to the middle third of the channel that is located between the dashed lines.}
\end{figure}
%\todo[inline]{Here is an example where you can avoid the passive voice because it is totally unnecessary. Instead of saying 'the bodies are eroded', simply say 'the bodies erode'. Also, a fluid cannot be Stokesian, since the Reynolds number depends on the length scale and velocity scale. It is the flow that is Stokes.}

@@

In this section, we describe an equivalent BIE formulation of the incompressible Stokes equations with Dirichlet boundary conditions
\begin{equation}
  \begin{split}
  \mu\Delta \uu &= \grad p\, , \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
  \end{split}
  \label{eqn:stokes}
\end{equation}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the eroding bodies $\gamma$.  In two dimensions, a BIE for a stream function formulation is possible~\cite{gre-kro-may1996}, but this does not extend to three dimensions.  Therefore, we adopt a primitive variable formulation where the velocity is written in terms of the standard double-layer potential~\cite{lad1963,poz1992}
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy\, , \quad \xx \in \Omega \, ,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown density function.  As the viscosity $\mu$ also sets the time scale, we take $\mu=1$.

@@

This BIE formulation requires that $\bd\Omega$ is smooth.  Unfortunately, smoothness is not guaranteed by erosion since corners develop at locations where the shear stress vanishes.  One option is to modify our BIE to handle corners~\cite{rac-ser2017, ser-rok2016, hel2011, gil-hao-mar2014, bre2012}, but these methods require the corner's location, a non-uniform spaced mesh, or are ill-conditioned.  Instead, we slightly modify the erosion model in Section~\ref{sec:thetaL} so that we can control the smoothness of the bodies.  In this manner, the BIE as described above can be applied.

%\todo[inline]{There is a typo in the above sentence (missing 'we'?). Also, technically there is  a dangling participle; i.e.~is it the erosion model which is in Section 2.3 or the modification? A slight rephrasing might clarify.}

@@

To numerically compute the completed double-layer potential~\eqref{eqn:completed_BIE}, we discretize each of the $M$ bodies with  $N_\iin$ points, and the outer boundary with $N_\out$ points.  Then, $N = MN_\iin + N_\out$ is the total number of discretization points of $\bd\Omega$.  The double-layer potential in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}\, , \quad i=1,\ldots,N \, ,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} \, ,
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is replaced with the limiting value of the double-layer potential
\begin{align}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)) \, ,
\end{align}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent vector at $\xx \in \bd\Omega$.  The equations for the Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are also discretized with the trapezoid rule.  Applying this discretization strategy to~\eqref{eqn:completed_BIE}, the dense linear system to be solved is
%\todo[inline]{Typo above}
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j) \, , \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j \, , \\ 
    \xi_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j \, ,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  Since all integrands are smooth and periodic, the trapezoid rule employed above achieves spectral accuracy~\cite{tre-wei2014}.

%We have applied the trapezoid rule since all the integrands are smooth and periodic, and this guarantees that the numerical solutions converges to the true solution with spectral accuracy~\cite{tre-wei2014}.
%\todo[inline]{The above sentence strikes me as awkward and verbose. Rephrase? \\
%How about this: Since all integrands are smooth and periodic, the trapezoid rule employed above achieves spectral accuracy.}

@@

We do not employ an FMM to evaluate the shear stress at $\xx_i$, $i=1,\ldots,N$, so the cost of evaluating the shear stress is $\bigO(N^2)$.  However, unlike the velocity double-layer potential that is evaluated at each GMRES iteration, the shear stress is only computed once per time step. In future work, we plan to use a kernel-independent FMM~\cite{yin-bir-zor2004} to evaluate the shear stress with $\bigO(N)$ operations.

%However, this calculation is only performed once per time step as opposed to the velocity double-layer potential, where we do employ the FMM, that must be evaluated at each GMRES iteration. 
%\todo[inline]{The above sentence is confusing. Try to rephrase and shorten it.}

@@

For the Stokes equations, on boundaries that have a no-slip boundary
condition, the shear stress is equivalent to the vorticity.  Therefore,
the vorticity is computed on the bodies $\gamma_\ell$ by numerically compute the shear stress as described in Section~\ref{sec:shearStress}.

@@

Naturally, the flow transmitted by the medium is linked to the amount of drag the bodies exert. We therefore show in Fig.~\ref{fig:mbodyplots}(c) the total drag exerted by all bodies as it varies in time. In both cases, the drag remains nearly constant throughout much of the simulation and then drops to zero in the final moments. Although all bodies are diminishing in size, the associated increase in $\umax$ leads to the nearly constant drag that is observed here.
